Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-03-11 submission/manuscript.Rmd
Rewrite introduction

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
source('../code/functions.R')

loadLibs(c("tidyverse", "knitr", "rmarkdown"))
```

```{r rdata_tables, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

# Read in meta data
metaf <- read_csv("../data/raw/metadata/good_metaf_final.csv")
metai <- read_csv("../data/raw/metadata/metaI_final.csv")
geof_samples <- read_tsv("../data/raw/metadata/303030_subset/NexteraXT003Map.tsv", 
                         col_names = F) %>% filter(X22 != "Negative")

# Read in analysis of SCFA by HPLC
hplc_groups_scfa_kruskal <- read_csv("../data/process/tables/acetate_kruskal_crc_groups.csv") %>% 
  mutate(scfa = "acetate") %>% slice(1) %>% 
  bind_rows(
    read_csv("../data/process/tables/butyrate_kruskal_crc_groups.csv") %>% 
      mutate(scfa = "butyrate") %>% slice(1), 
    read_csv("../data/process/tables/propionate_kruskal_crc_groups.csv") %>% 
      mutate(scfa = "propionate") %>% slice(1)) %>% 
  mutate(bh = p.adjust(pvalue, method = "BH"))

hplc_treat_scfa_kruskal <- read_csv("../data/process/tables/scfa_treatment_comparison_pvalues.csv") %>% 
  filter(scfa != "isobutyrate")

# Load in Random Forest with and without SCFA summary data
disease_rf <- read_csv("../data/process/tables/rf_model_ttest_comparisons.csv")

# Load in data for select genes in SCFA pathways
picrust_select <- read_csv("../data/process/tables/specific_scfa_kruskal_picrust_summary.csv")
picrust_select_treat <- read_csv("../data/process/tables/specific_scfa_wilcox_picrust_treatment_summary.csv")
opf_select <- read_csv("../data/process/tables/select_scfa_opf_kruskal_summary.csv")

# Load in the group comparison from SCFA models
class_group_rf_data <- read_csv("../data/process/tables/full_model_class_group_results.csv")
reg_group_rf_data <- read_csv("../data/process/tables/full_model_reg_group_results.csv")

# Load in sig associations and top 10 RF model presence
sig_assoc_top10_model <- read.csv("../data/process/tables/sig_association_and_RF_top10.csv")

```

@@

Running title: SCFAs and colorectal tumors

@@

Marc A. Sze${^1}$, Nicholas A. Lesniak${^1}$, Mack T. Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$

@@

**Background.** Colorectal cancer (CRC) is increasing in prevalence in individuals under 50 and because of this will be a continuing health concern for the foreseeable future. The majority of the risk for developing CRC is attributable to environmental factors. One of these environmental factors is the microbiota, with certain bacterial community members being more prevalent in CRC while other taxa are higher in individuals without tumors. Some of the bacterial species in taxa that are more abundant in individuals without tumors can use fiber to produce short-chain fatty acids (SCFAs) that inhibit tumor growth in model systems. However, the data supporting the importance of SCFAs in human CRC is less certain. Here, we test the hypothesis that SCFA concentrations and taxa associated with their production are different in individuals with colorectal tumors.

**Methods.** We analyzed a cross-sectional (n=`r length(metai$sample)`) and longitudinal pre- and post-treatment (n=`r length(metaf$initial)`) group for their fecal concentrations of acetate, butyrate, and propionate. Analysis also included imputed gene relative abundance with PICRUSt, metagenomic sequencing on a subset (n=`r length(geof_samples$X30)`) of the total cross-sectional group, and tumor classification and SCFA prediction models using Random Forest.

**Results.** No difference in SCFA concentrations were found between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15). Using metagenomic sequencing, there was also no difference in genes involved with SCFA synthesis between individuals without tumors and patients with adenomas or carcinomas (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70). Finally, there was no difference between the ability of Random Forest models to predict patients with adenomas or carcinomas versus individuals without tumors (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05).

**Conclusions.** Although our data does not support the hypothesis that fecal SCFA concentrations are different in the general CRC population, there still may be specific types of colorectal tumors where SCFAs may be beneficial for treatment of CRC. Alternatively, our observations also support the hypothesis that there may be other metabolites or mechanisms (e.g. bacterial niche exclusion) that may be more protective against tumorigenesis and have not been thoroughly investigated in the context of human CRC.

@@

## Introduction

Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US and the prevelance is increasing in invidiauls under 50 years of age [@Haggar2009; @Siegel2016]. Although there is a genetic component to the disease, the environment is considered a larger risk factor for CRC [@Lichtenstein2000]. These environmental risk factors include but are not limited to smoking cigarettes, diet, and the microbiota [@FlissIsakov2017; @Lee2015; @Kostic2011]. Many of these environmental risk factors, including the microbiota, are modifiable. This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered in CRC [@Shah2017; @meta_analysis_crc_Sze2018]. Many of these previous case/control studies have identified resident bacterial taxa to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @meta_analysis_crc_Sze2018]. Many of the bacterial species from these resident taxa identified in these case/control studies actively produce short-chain fatty acids (SCFAs) from fiber that are part of our general diet [@Holscher2017]. The most extensively studied SCFAs are acetate, butyrate, and propionate [@Louis2016]. These SCFAs are hypothesized to be the main metabolites involved with protection against tumorigenesis and could help to reduce the risk of CRC.

Prior research suggests that SCFAs have promise in acting as an anti-tumorigenic agent. Specific SCFAs have shown positive results within model systems [@test_OKeefe2016]. For example, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018]. Additionally, fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018]. These exciting results in model systems suggest that supplementation with food sources that bacteria use to create these SCFAs may confer beneficial effects against CRC. However, it is important to note that these model systems provide only preliminary evidence towards the ability of SCFAs to reduce and treat tumors and the studies reporting benefit in humans has been less convincing.

Overall, there is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations. The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus patients with adenomas or individuals without colon tumors [@Ohigashi2013]. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017]. These findings argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis. Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC, there is a need for more investigation into this area.

Our study fills some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations. Specifically, by using a seperate cohort it tests previous case/control findings on SCFA concentrations in individuals with and without tumors. Additionally, prior investigations grouped patients with adenoma and individuals without tumors into a single group to compare against patients with carcinomas. Despite doing this, the suggestion was made that a reduction in SCFA concentration would also be observed between patients with adenomas and indivdiuals without colon tumors. We also can test this suggestion because we have a larger number of patients with adenomas within our study and do not need to group patients with adenomas and individuals without colon tumors into a single group. Additionally, we build upon these observations by assessing the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a risk stratification tool of colorectal tumors (adenoma or carcinoma). We also investigate whether OTUs that are most important to these models are closely associated with the classification of SCFA concentrations. Collectively, this study provides important information on the replicability of previous findings in humans by extensively studying how SCFAs are associated with colorectal tumors.

To accomplish this task we directly measured the concentration of acetate, butyrate, and propionate within fecal samples for two different groups. The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors. To provide further support for our SCFA findings we also used PICRUSt [@Langille2013] and metagenomic sequencing to investigate differences in genes encoding enzymes involved with SCFA synthesis between individuals without colon tumors, patients with adenomas, and patients with carcinomas. Next, we investigated whether taxa associated with SCFA production were important to disease classification models. First, using the cross-sectional data, we analyzed the number of correlations between OTU relative abundance and SCFA concentrations across individuals without colon tumors and patients with adenomas or carcinomas. Second, we assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenomas or carcinomas using the Random Forest algorithm [@randomforest_citation_2002]. Third, we analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations. Finally, we compared whether SCFA concentrations replace important taxa in disease models and whether the same OTUs are the most important variables used to classify disease and SCFA concentration. Collectively, this investigation provides additional information as to whether SCFAs are decreased in patients with colorectal tumors and provides context as to whether targeting taxa to increase SCFA concentrations is a viable option to protect against colon tumorigenesis.

@@

## Results

**Decreased SCFA concentrations are not associated with adenoma or carcinoma tumors.** We used high-performance liquid chromatography (HPLC) to measure acetate, butyrate, and propionate concentrations of frozen fecal samples from `r length(metai$sample)` individuals at a cross-sectional point in time. There was no difference between individuals without colon tumors (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C]. We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples. Although there was a general trend for increasing acetate, butyrate, and propionate concentrations following treatment for tumors, there was no significant difference between pre- and post-treatment for either patients with adenomas (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinomas (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].

**Gene abundance for enzymes involved in SCFA synthesis are the same for individuals without colon tumors and patients with adenomas or carcinomas.** In order to provide further confirmation and support of our SCFA concentration results we investigated the genes encoding specific enzymes involved with SCFA synthesis [Table S1]. Using this list of specific genes [Table S1], we looked for differences in gene abundance between individuals without colon tumors and patients with adenomas or carcinomas. Although we intended to analyze all the genes in the list [Table S1], not all of the KEGG genes were identified during our analysis. We first analyzed imputed gene relative abundance, calculated from our OTU data that was generated from 16S rRNA gene sequencing. We found no difference in any of the imputed gene relative abundances for enzymes involved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2]. Since butyrate is one of the most studied SCFAs in the context of CRC, we visualized the observed lack of difference between indviduals without colon tumors and patients with adenomas or carcinomas in gene abundance for enzymes involved with SCFA synthesis using the butyrate kinase gene [Figure 2A]. Additionally, using a paired Wilcoxon rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3]. Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results. Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals without colon tumors (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenomas (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinomas (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4]. This similarity between indviduals without colon tumors and patients with adenomas or carcinomas is highlighted again by visualizing the gene abundance for the butyrate kinase gene [Figure 2B]. These observations provide evidence that gene prevalence for enzymes involved in SCFA synthesis does not change due to colorectal tumors and provides further support for our original SCFA concentration observations.

**The number of OTUs positively associated with SCFA concentration were similar between individuals without colon tumors and patients with adenomas or carcinomas.** Having found no difference between individuals without colon tumors and patients with adenomas or carcinomas in SCFA concentrations or genes encoding enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations. The main goal of this analysis was to identify if there were OTUs that were significantly associated with SCFA concentrations and if this was different between individuals that did not have colon tumors, had an adenoma, or had a carcinoma. To accomplish this we used Spearman's rho, a non-parametric measure of association, and tested if there was a correlation that was signficantly greater than zero. We found that taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* dominated statistically significant OTU correlations [Figure 3 & Table S5]. There was a noticeably higher number of significant negative correlations associated with patients with adenomas for all SCFAs tested [Figure 3]. In particular, OTUs from the *Ruminococcaceae* family had the largest share of these negative correlations within patients with adenomas [Figure 3]. Patients with adenomas also had more positive correlations between OTUs and SCFA concentrations, but their total number was more similar to individuals without colon tumors or patients with carcinomas versus the analogous comparison for the number of negative correlations [Figure 3]. Additionally, the number of positive correlations between OTUs and SCFA concentrations was similar between individuals without colon tumors and patients with a carcinoma [Figure 3]. Finally, when we used high/low SCFA groups based on the overall median concentration for each SCFA instead of SCFA concentrations a similar pattern was still observed [Figure S1 & Table S6]. Overall, these results suggest that the resident taxa that may change the most due to colon tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.

**SCFA concentrations do not replace important *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* OTUs in Random Forest models built to classify tumors.** Despite the lack of difference in positive correlations between OTUs and SCFA concentrations between individuals with and without tumors, OTUs associated with SCFA concentrations could still be the most important variables to Random Forest models built to classify patients with adenomas or carcinomas. We tested this by using the Random Forest algorithm to build models with OTU abundance data or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples. With these models we compared whether OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained when SCFA concentrations were added to the model. Additionally, we also compared whether any of the important OTUs that remained also had significant correlations with SCFA concentrations. Both our adenoma and carcinoma models classified patients with a similar degree of success, as measured by the area under the curve (AUC) (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4A & 4D]. After the addition of SCFA concentrations to the adenoma or carcinoma models, many OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained as important variables to the model [Figure Figure 4B-C & 4E-F]. After adding SCFA concentrations to the adenoma model, there were only `r length(unique(sig_assoc_top10_model %>% filter(model == "adn_full") %>% pull(otu)))` OTUs significantly associated with SCFA concentration that were part of the top 10 most important variables and both were postively associated with acetate, butyrate, or propionate concentrations. When SCFA concentrations were added to the carcinoma model, only `r length(unique(sig_assoc_top10_model %>% filter(model == "crc_full") %>% pull(otu)))` OTU that was associated with SCFA concentration was in the top 10 most important variables and it was negatively associated with acetate and butyrate concentrations. In combination with the previous results on OTU correlations, these observations provide additional evidence that the resident taxa that are associated with protection against tumorigeneis are not ones associated with acetate, butyrate, or propionate production.

**The most important OTUs for SCFA concentration and tumor Random Forest models are not the same.** It is possible that due to the way the Random Forest algorithm works, OTUs associated with SCFA concentrations could be downweighted in importance within the adenoma or carcinoma models when SCFA concentrations are included. To test if this is the case we used OTU data and built Random Forest models to classify SCFA concentrations. Overall, the correlation between the predicted and actual SCFA concentrations were moderately associated with each other [Figure 5A]. Additionally, because the training set R^2^ was always higher than the test set R^2^, all SCFA concentration models tended to be over fit, suggesting that rarer taxa were important for these classifications [Figure 5A]. There also was a difference in accuracy based on whether the fecal sample was from an individual without colon tumors or from patients with adenomas or carcinomas [Figure 5B]. When comparing the adenoma or carcinoma model to the SCFA concentration models there was minimal overlap between these model's most important OTUs [Figure 4B-C, 4E-F and 5C-E]. The only OTU that was in the top 10 most important variables and had overlap between the models was OTU00167 (*Clostridiales*) [Figure 4B-C, 4E-F, 5C-E]. Similar observations were made when using high/low SCFA groups based on the median SCFA concentration [Figure S2]. Collectively, these observations provide evidence that it is possible to identify specific OTUs associated with higher SCFA concentrations and accordingly these OTUs belong to taxa known to produce acetate, butyrate, and propionate. Although it is possible to identify OTUs associated with SCFA production, overall, our results do not support the hypothesis that SCFA concentration or OTUs associated with their production are different between individuals with no colon tumors and patients with adenomas or carcinomas.

@@

## Discussion

The observations from this study do not support the hypothesis that SCFA concentrations are different in individuals with tumors. Whether we directly measured the SCFA concentration or investigated genes that encoded enzymes used for their production, no difference could be identified between individuals without tumors and patients with adenomas or carcinomas [Figure 1 & 2]. There were differences in the number of significant correlations between SCFA concentration and OTU relative abundance based on whether individuals did not have colon tumors, had an adenoma, or had a carcinoma. However, SCFA concentrations did not provide increased model accuracy for tumor classification [Figure 3-4 & S1]. In models with SCFA concentrations included, many OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* remained as important variables for the model [Figure 4]. Additionally, when models using OTU relative abundance to classify SCFA concentrations were assessed, the OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* were not the same as the OTUs that classified to these taxa in the tumor models [Figure 4-5]. Collectively, our observations suggest that the resident taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* that are different between individuals without colon tumors and patients with adenomas or carcinomas, are not the same as those involved with SCFA production.

Although SCFAs have been shown to be anti-tumorigenic, most of these studies have been performed in model systems [@test_OKeefe2016; @Encarnao2018]. Many of the *in vivo* studies use proxies such as fiber supplementation rather than SCFAs directly [@Holscher2017]. Although it is well known that breakdown products from gut bacteria results in SCFA production, fiber effects on tumorigenesis may be through other mechanisms in these *in vivo* model systems. Additionally, the observations in humans on the benefit of SCFAs in preventing tumorigenesis have been mixed. In previous case/control studies lower SCFA concentrations were observed in patients with carcinomas versus those without carcinomas [@Ohigashi2013]. Yet, this is in contrast to multiple randomized-controlled trials that have found no difference in tumor recurrence between patients who do and do not get fiber supplementation [@Schatzkin2000; @Yao2017]. In contrast to the *in vivo* model findings, the observations made in these randomized-controlled trials would suggest that SCFAs do not prevent or slow tumorigenesis. One reason for these results is that SCFA concentrations and responses to fiber vary quite a bit between healthy individuals [@scfa_measures_venkataraman2016]. This information taken together with our observations would suggest that either SCFAs provide little to no benefit as an anti-tumorigenic compound in colorectal cancer or that individuals who do not respond to fiber supplementation would need to acquire these bacteria to achieve a benefit.

Another possible alternative explanation as to why no difference in SCFA concentration between individuals without colon tumors and patients with adenomas or carcinomas was observed, could be because only certain types of colorectal cancers are affected by SCFAs. One limitation of current research into the effect of SCFAs and the microbiota in CRC has been that all tumors are treated as the same type. However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in certain subsets of individuals. Similar to the idea of using immunotherapy as a targeted treatment option for specific tumors [@Thomas2016], SCFAs may have beneficial effects for distinct types of colorectal tumors. Future research will need to test if this is a valid hypothesis. Regardless of this limitation, our results in combination to previous randomized controlled trials on fiber supplementation [@Schatzkin2000; @Yao2017] suggests that using SCFAs as a general treatment for colorectal cancer is unlikely to provide a reduction in tumorigenesis.

One possible technical limitation is that a fecal sample may not be an ideal type of bio-specimen and that the effect SCFAs have on tumorigenesis is only detected in the colon. However, this is unlikely to be a major confounder. First, most *in vivo* studies, as well as human studies, have used fecal material in their analysis [@Ohigashi2013; @Bishehsari2018]. Second, previous studies that measure SCFA changes after fiber supplementation use fecal material to track these responses with a great deal of success [@scfa_measures_venkataraman2016]. Although there are limitations with the current research on SCFAs and colorectal tumors, technical limitations are less likely to be the cause of this. Clearly, our observations provide robust evidence that these specific metabolites may not be protective or used as a general treatment option in colorectal cancer. Yet, taxa that are associated with SCFA production are consistently higher in individuals without colon tumors than patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].

The potential protection against colorectal cancers may not be from SCFAs, even though taxa associated with their production are higher in individuals without tumors versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018]. Our data is consistent with the literature despite finding no difference in SCFA concentrations between individuals without colon tumors and patients with adenomas or carcinomas. In particular, we observed that the important OTUs in our tumor models were to taxa that are associated with SCFA production but that these OTUs were not themselves associated with acetate, butyrate, or propionate production. One of the clearest results to highlight this within our study were the results showing that different OTUs from the same taxonomic classification are in tumor and SCFA Random Forest models [Figure 4 & 5]. This leads to the possibility that protection may be through other routes. For example, there could be a different pathway or other less extensively studied metabolites that provide the necessary protection against tumorigenesis. Alternatively, previous research has shown that non-resident oral microbes, such as *Fusobacterium*, *Porphyromonas*, *Parvimonas*, and *Peptostreptococcus* are associated with carcinomas [@Zeller2014; @Baxter2016; @Kostic2011] and can increase inflammation [@Flynn2016]. When these non-resident oral microbes increase inflammation they change the existing environment to one in which is more hospitable for themselves [@Flynn2016]. It is possible that these resident microbes provide protection not through a metabolite but instead through active exclusion of these non-resident oral microbes which in turn prevents them from establishing themselves within the community. Overall, even though we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that our results open up many new and exciting avenues for investigation into how metabolites and the microbiota can prevent tumorigenesis.

## Conclusions

Our observations found no difference in SCFA concentration or in genes encoding enzymes involved in their production between individuals without colon tumors and patients with either adenoma or carcinoma tumors. The SCFA concentration also did not improve classification of adenoma or carcinoma tumors. The most important OTUs to these models classifying patients with adenomas or carcinomas were mostly not associated with SCFA production. Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence. Additionally, these observations suggest that resident microbes that are not involved in SCFA production may be the important resident community members involved with preventing tumorigenesis. By focusing on alternative mechanisms that are associated with these non-SCFA producing resident microbes or SCFA effects on distinct colorectal cancer groups the identification of more promising therapeutic options for use in treating colorectal cancer may be found.

@@

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =`r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())` and carcinoma n = `r length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor (surgery or colonoscopy) with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.

@@

**16s rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017]. In brief, sequences were quality filtered and contigs created from the paired end reads. Any sequences with ambiguous base calls were discarded. Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012]. Chimeras were identified and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptiClust algorithm [@opticlust_westcott2017].

>>>>

@@

Fecal short chain fatty acids are not predictive of colorectal cancer status and cannot be predicted based on bacterial community structure

@@

Marc A. Sze${^1}$, Begüm Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Mack T. Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$

@@

@@

## Introduction (400 words)

Colorectal cancer is the third leading cancer-related cause of death within the United States [@Haggar2009; @Siegel2016]. Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990]. This leaves a significant role for environmental,  behavioral, and other factors such as smoking and diet [@FlissIsakov2017; @Lee2015]. Colorectal cancer is thought to be initiated by a series of mutations that accumulate as mutated cells proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990]. Throughout this progression, there are ample opportunities for bacterial populations to create mutations, induce inflammation, and accelerate tumorigenesis [@Flynn2016]. Numerous studies in murine models have supported this model [@Zackular2013; @Baxter2014; @Zackular2015; @DeStefanoShields2016; @Tomkovich2017]. Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Kostic2011; @Sze2018; @Baxter2016; @Zackular2014; @Zeller2014; @Shah2017; @Dejea2018; @Arthur2012]. These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.

SCFAs have have anti-inflammatory and anti-proliferative activities [@Encarnao2018; @Verma2018; @Zheng2017; @OKeefe2016]. Furthermore, manipulation of SCFAs in mouse models of colorectal cancer by direct supplementation or feeding of fiber caused an overall reduction in tumor burden [@Bishehsari2018; @Tian2018]. These results suggest that supplementation with substrates that bacteria can ferment to produce SCFAs may confer beneficial effects against colorectal cancer. Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans. Based on similar observations, many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Vital2014; @Sanna2019; @Liu2019; @Meisel2016]. Case-control studies that have investigated SCFA concentrations in colorectal cancer found that patients with carcinomas had lower concentrations of SCFAs versus patients with adenomas or individuals without colon tumors [@Ohigashi2013]. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, in randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Schatzkin2000; @Yao2017; @Kunzmann2015; @Murphy2012; @Gianfredi2018]. These findings temper enthusiasm for treatments that aim to use SCFAs as biomarkers or protection against tumorigenesis.

To better understand the connection between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of SCFAs in feces of previously characterized individuals with normal colons, adenomas, and carcinomas and subset of those individuals after they underwent treatment for their lesions. Our goals included: (1) testing whether there was an association between SCFA concentration and tumor status, (2) determining whether SCFA concentrations could be used as biomarkers to improve the detection of colonic lesions, and (3) predicting SCFA concentrations based on the relative abundance of fecal bacteria and their genes.

@@

## Results (700 words)

* There is no association between presence of tumors and SCFA concentration
	- Wilcoxon/T-test
	- Cross sectional
	- Aggregating SCFAs does not improve association with CRC
	- Combining 16S and SCFA does not improve prediction of CRC
	- Lack of change pre and post treatment
* Community structure is not predictive of SCFA production
	- Random forest regression
	- 16S
	- picrust
	- shotgun metagenomic data

@@

## Discussion (200 words)

* Unlikely that SCFAs are the primary mechanism that limits tumorigenesis
* Limitations
	- Observing the community once the tumor is developed
	- Not looking at mucosal communities
* Need to identify other possible mechanisms that drive and prevent tumorigenesis
* Supports evidence that fiber consumption has a null effect on tumorigenesis

@@

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @colorectal cancer_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =`r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())` and carcinoma n = `r length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor (surgery or colonoscopy) with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @colorectal cancer_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.

@@

**16S rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017]. In brief, sequences were quality filtered and contigs created from the paired end reads. Any sequences with ambiguous base calls were discarded. Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012]. Chimeras were identified and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptiClust algorithm [@opticlust_westcott2017].
